You have more money in your wallet. Does that mean you will spend extra on food?

### 2024-04-06 paper/paper.qmd
finished the discussion section

>>>>

## Factoring in elasticities {#sec-ela}

Income and price elasticities of food is another factor that explains why low change in food expenditure for change in income. Income elasticity considers how sensitive are demand for food when income changes, while price elasticies looks at how change in price affects demand of food. @almaas2019income considers an experimental study in Kenya that randomized cash transfers to find income elasticity of food to 0.87. @purcell1967quantity expands on this and finds that income elasticies for items like poultry, fish and seafood, fruits were below 0.5. This establishes that food is normal commodity and as income rises, quantity demand for food also rises, albeit marginally. Further, @andreyeva2010impact conducted a meta-analysis on price elasticities of different food categories and found that essential times including fish, beef, poultry, fruits and vegetables ranged below 0.75. This shows that when price of food increases, demand for it goes down only slightly. This constitutes why we see only 7% economic significance as demand for food moderately increases when income goes and moderately decreases when prices of food goes up. Demand for food does not experience major shift due to changes in income.

## Weaknesses and next steps

The weaknesses of our model derives from the data we used. We used quarterly data from 2007 and 2022. We chose this range of dates because we were interested how the relationship between food expenditure and income changed over this time. This time period also captured the economic swings of Covid shocks. US and much of the world went into the pandemic in the fall of 2020. During the pandemic period, disposable income went up  sharply (see @fig-1) because social benefit programs put additional wealth into consumers pockets and moreover, consumers were not able to spend this wealth due to lockdown measures. Food expenditure, however, experienced a more diminished shock. This is because food is an essential commodity and one that you need for survival. Historical data on food elasticies also predict why expenditure on food does not change much as discussed on @sec-ela. This resulted in data where we see sharp and unpredictable swings in disposable income and predictable change in food expenditure and this distorted our data.

This distortion is a problem this causes model to inaccurately estimate estimand. The true underlying relationship may not be as accurate as we would have hoped because the model yields biased estimates. It also diminishes external validity. Global pandemic is an unique situation and it introduced unique shocks that we do not foresee happening on a regular basis. This diminishes the generalizability and applicability of the model under normal circumstances. Lastly, policies around social benefit measures different from country to country during pandemic and further, lockdown measures differed between countries as well. It is plausible that economic swings US experienced may be different to that of other countries. Again, this reduces external validity. If we consider another country in the same 15 year period, our model may not be applicable as Covid-related policies were not the same as US's.

In the next iteration of the paper, I would consider a time frame that excludes large scale shocks on global scale. As seen with Covid, it diminishes external validity and obscures estimand, which is what we are after.
